---
title: "pathway_analysis"
author: "Emily Carroll"
date: '2022-10-21'
output:
  pdf_document: default
  html_document: default
---
This is the pathway analysis practical of the OBDS course 
- The first part of this pracical is using gprofiler to perform ORA on output dataset of differentially expressed genes from DESeq2


# load packages 
```{r}
# load librarys

library(tidyverse)
library(gprofiler2)
```

# Read in our tables from this morning in a new Rmd file
```{r}
# read in files from this morning 

# 1. all results (this was filtered to remove non DE genes in our DESeq2 list so only contains genes expressed in our samples) -> call this table 'all_genes_df'  - this will be our background

all_genes_df <- read_csv("/project/obds/ecarroll/DESeq2_tutorial/CD8_vs_CD4_DE_results_lfcshrinkage_allgenes.csv")
head(all_genes_df)

#filter out the genes which padj is NA 

all_genes_df_filtered <- all_genes_df %>% 
  dplyr::filter(!is.na(padj))

# 2. all significant upregulated genes -> call this upreg_genes_df

upreg_genes_df <- read_csv("/project/obds/ecarroll/DESeq2_tutorial/CD8_vs_CD4_DKO_DE_results_lfcshrinkage_0.05_log2FC1_up.csv")
head(upreg_genes_df)

# 3. all significant down regulated genes -> call this downreg_genes_df 

downreg_genes_df <- read_csv("/project/obds/ecarroll/DESeq2_tutorial/CD8_vs_CD4_DKO_DE_results_lfcshrinkage_0.05_log2FC1_down.csv")
head(downreg_genes_df)

# Check the dimensions of each of your dataframes using dim()
dim(all_genes_df)
dim(all_genes_df_filtered)
dim(downreg_genes_df)
dim(upreg_genes_df)

# upreg_genes_df should be 915
# downreg_genes_df should be 1204
# all_genes_df_filtered should be 17729


```
Now we want a vector of gene ids to use for analysis 

```{r}
# from each of these dataframes get a vector of the ensembl ids 
all_gene_ids <- all_genes_df_filtered$ensembl_gene_id
    
upreg_gene_ids <- upreg_genes_df$ensembl_gene_id
    
downreg_gene_ids <- downreg_genes_df$ensembl_gene_id


```

#Perform ORA using g:profiler 

1. Look at documentation - https://cran.r-project.org/web/packages/gprofiler2/vignettes/gprofiler2.html
```{r}
# use the gost() function - what paramters should we include/change 
# we want to test the ORA of upreg_gene_ids vs our all_gene_ids

# get help for function
?gost() 

# write query and assign to output variable named `gost_results_obj`
# multi-query could give up reg and down reg genes at the same time, but we don't want to do that right now. If not don't need to include as the default is false.
# exclude_iea change to TRUE (default is FALSE)
# user threshold change to 0.01 - default is 0.05
# custom_bg: background - need to put your background in here - all_gene_ids (includes all genes (upregulated, downregulated and non-significant))

gost_results_obj <- gost(query = upreg_gene_ids, 
                organism = "mmusculus", 
                ordered_query = FALSE, 
                multi_query = FALSE, 
                significant = TRUE, 
                exclude_iea = TRUE, 
                measure_underrepresentation = FALSE, 
                evcodes = TRUE, 
                user_threshold = 0.01, 
                correction_method = "g_SCS", 
                custom_bg = all_gene_ids, 
                numeric_ns = "", 
                sources = NULL,
                as_short_link = FALSE)

# what are our outputs  
gost_results_obj

```

```{r}
# plot our results using in built function in gprofiler -> which one is it? 
# capped = TRUE highlights more significant stuff but also keeps more specific stuff

gostplot(gost_results_obj, capped = TRUE, interactive = TRUE)

```

```{r}
# what does as_short_link = FALSE do in gost() function? 
# indicator to return results as short-link to the g:Profiler web tool. If set to TRUE, then the function returns the results URL as a character string instead of the data.frame.

gost_results_obj_new <- gost(query = upreg_gene_ids, 
                organism = "mmusculus", 
                ordered_query = FALSE, 
                multi_query = FALSE, 
                significant = TRUE, 
                exclude_iea = TRUE, 
                measure_underrepresentation = FALSE, 
                evcodes = TRUE, 
                user_threshold = 0.01, 
                correction_method = "g_SCS", 
                custom_bg = all_gene_ids, 
                numeric_ns = "", 
                sources = NULL,
                as_short_link = TRUE)

# then when we look at the object it gives us a https link, which if we look at it in a browser it gives us our plot and this plot is interactive - when you scroll over the dots it gives you the GO term and the name of the pathway (?)
# it also gives you detailed results
# if you limit the range in the detailed results to 300-500 it gives you a lot more manageable and specific results
# the link does expire but you can download the plots in the meantime 

gost_results_obj_new


```

Answer -> This is the link it returns: "https://biit.cs.ut.ee/gplink/l/ruq8XOltRS"

```{r}
# Filter results table for genesets with between 5-300 genes 

#genesets_filtered <- gost_results_obj %>% dplyr::filter()


```


2. Try querying multiple gene lists at the same time (up and down regulated genes) using Multiple queries - see the documentation 

```{r}

```

Can you plot these querys together?

```{r}


```

Output the get_version_info() to save the versions of the databases that were used for the analysis!! 
```{r}
get_version_info((organism="mmusculus"))

```


```{r}
# This tells you what version of R you used, what version - good to use for paper publication - knit R markdownfile

sessionInfo()


```

