---
title: "Example code for ggplot2"
author: "Kevin Rue-Albrecht"
date: "03/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(patchwork)
library(gridExtra)
```

# Exercise

Use `library(ggplot2)` to load the package.

```{r}
library(ggplot2)
```
```{r}
head(diamonds) #use head to check the dataset
```

## Specify the dataset.

What does `ggplot(diamonds)` do? It's blank because it doesn't know what to do as you haven't specified anything. 

```{r}
ggplot(diamonds) 
```

> Answer:

## Add the aesthetics.

What does `ggplot(diamonds, aes(x = carat, y = price))` do? Knows what goes on x and y but doesnt know how to plot it as we haven't specified a graph type.

```{r}
ggplot(diamonds, aes(x = carat, y = price))
```

> Answer:

## Add geometric objects

- Add data points showing `carat` on the x-axis and `price` on the y-axis.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
    geom_point()

  
```

- Color data points by `cut`.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
    geom_point(colour = "blue", size = 0.7) # simple way of doing it to colour every dot the same
  
```
```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
    geom_point(aes(colour = cut), size = 0.7) # using "cut" function to add colours based on the quality of the cut
```

- Add a smoothed mean trend line.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
    geom_point(aes(colour = cut), size = 0.7) +
    geom_smooth() # use geom_smooth to add a trend line

```

- Assign that last plot to an object called `obds_diamonds`.

```{r}

obds_diamonds <- ggplot(diamonds, aes(x = carat, y = price)) +
    geom_point(aes(colour = cut), size = 0.7) +
    geom_smooth() # assign it to obds_diamonds 

```


# Exercise

Predict the difference between these two plots

### Plot 1

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point() +
  geom_smooth() # as the colour=cut function is in the ggplot statement it also colours the smooth lines by cut - so you have 5 trend lines coloured by cut compared to the above plot which only has one trend line
```

### Plot 2

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(colour = cut)) +
  geom_smooth()
```

> Answer:

# Exercise

## Trend lines

Using the `ChickWeight` data set:

```{r}
head(ChickWeight) # see what the data looks like
```


- Create a scatter plot of weight (y-axis) over time (x-axis).

```{r}
ggplot(ChickWeight, aes(x=Time, y=weight)) +
  geom_point()

```

- Color by diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x=Time, y=weight)) +
  geom_point(aes(colour=Diet)) # note that R plots all of 1 first, then all of 2, then 3 and 4. This can be misleading as it will plot the last values on the top. Note here 4 (purple) is clearer colour.

```

- Add a linear mean trend line for each diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x=Time, y=weight, colour=Diet)) +
  geom_point() +
  geom_smooth(method=lm) # method = lm is for linear model

```

At this point you should be able to visually identify
the diet that leads to the highest mean increase in weight.

> Answer:

- Facet a ribbon of sub-plots, one per diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x=Time, y=weight, colour=Diet)) +
  geom_point() +
  geom_smooth(method=lm) +
  facet_wrap(~Diet, nrow=2)
```

- Assign that last plot to an object called `obds_chickweight`.

```{r}

obds_chickweight <- ggplot(ChickWeight, aes(x=Time, y=weight, colour=Diet)) +
  geom_point() +
  geom_smooth(method=lm) +
  facet_wrap(~Diet, nrow=2)
```

# Exercise

## Bar plot

- Load the `ggplot2::msleep` data set.

```{r}
msleep

```

- Draw a bar plot of number of observations (i.e., rows) for each taxonomic order
  (i.e, one plot and one bar per taxonomic order).

```{r}
ggplot(msleep, aes(x=order)) +
  geom_bar()

```

- Change the angle and font size of the text for the x-axis ticks (not the axis titles).
  Justify the text of those x-axis ticks as right-aligned.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(msleep, aes(x=order)) +
  geom_bar(aes(fill=order)) +
  theme(axis.text.x=element_text(angle=45, size=rel(0.9), hjust=1, vjust=1))

# rel(0.9) means make the size of the text 0.9 relative to the original text. hjust=1 means right align

```

- Change the value and font size of the title for both x and y axes.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x=order)) +
geom_bar(aes(fill=order)) +
theme(axis.text.x = element_text(angle=45,
                                 size=rel(0.7),
                                 hjust=1, vjust=1),
                                 axis.title = element_text(size=rel(1.2))) +
labs(x= "Order of Mammals", y="Number of Observations")

```

- Fill each bar with colors, proportionally to the count of each genus.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(msleep, aes(x=order)) +
geom_bar(aes(fill=genus)) + #change fill=genus
theme(axis.text.x = element_text(angle=45,
                                 size=rel(0.7),
                                 hjust=1, vjust=1),
                                 axis.title = element_text(size=rel(1.2))) +
labs(x= "Order of Mammals", y="Number of Observations")

```

From this point onwards, you may need to iteratively resize the text of the ticks and axes for readability.

- Reduce the legend key size.
  (Recommendation: use `unit(2, "mm")`).
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x=order)) +
geom_bar(aes(fill=genus)) +
theme(axis.text.x = element_text(angle=45,
                                 size=rel(0.7),
                                 hjust=1, vjust=1),
                                 axis.title = element_text(size=rel(1.2)),
                                 legend.key.size=unit(2, "mm"), #specify legend key size to 2mm
                                 legend.text=element_text(size=rel(0.5)) 
                   ) +
labs(x= "Order of Mammals", y="Number of Observations")

```

- Force the legend to be display in 3 columns.
  (Recommendation: use `guide_legend(...)`).
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(msleep, aes(x=order)) +
geom_bar(aes(fill=genus)) +
theme(axis.text.x = element_text(angle=45,
                                 size=rel(0.7),
                                 hjust=1, vjust=1),
                                 axis.title = element_text(size=rel(1.2)),
                                 legend.key.size=unit(2, "mm"), #specify legend key size to 2mm
                                 legend.text=element_text(size=rel(0.5))
                   ) +
labs(x= "Order of Mammals", y="Number of Observations") +
guides(fill=guide_legend(ncol=3))
       
```

- Add a contour of thin black lines to the bars.

```{r}

ggplot(msleep, aes(x=order)) +
geom_bar(aes(fill=genus), colour="black", size=0.1) + #added colour="black" and size=0.1 to make a black line with 0.1 size around each of the bars
theme(axis.text.x = element_text(angle=45,
                                 size=rel(0.7),
                                 hjust=1, vjust=1),
                                 axis.title = element_text(size=rel(1.2)),
                                 legend.key.size=unit(2, "mm"), #specify legend key size to 2mm
                                 legend.text=element_text(size=rel(0.5))
                   ) +
labs(x= "Order of Mammals", y="Number of Observations") +
guides(fill=guide_legend(ncol=3))

```

- Assign that last plot to an object called `obds_msleep`.

```{r}

obds_msleep <- ggplot(msleep, aes(x=order)) +
geom_bar(aes(fill=genus), colour="black", size=0.1) + 
theme(axis.text.x = element_text(angle=45,
                                 size=rel(0.7),
                                 hjust=1, vjust=1),
                                 axis.title = element_text(size=rel(1.2)),
                                 legend.key.size=unit(2, "mm"), #specify legend key size to 2mm
                                 legend.text=element_text(size=rel(0.5))
                   ) +
labs(x= "Order of Mammals", y="Number of Observations") +
guides(fill=guide_legend(ncol=3))

```

# Exercise

## Plotting grid

- Collate the plots that we assigned to objects through the day, as a single plot.

  + Plots: `obds_diamonds`, `obds_chickweight`, `obds_msleep`.
  
  + Methods: `cowplot::plot_grid()`, `patchwork`, `gridExtra::grid.arrange()`.

Using `cowplot`.

```{r, fig.height=6, fig.width=6}
library(cowplot)
                   
first_row <- cowplot::plot_grid(obds_diamonds, obds_chickweight, labels=c("A", "B"), ncol=2, nrow=1) #use the first row and stitch together the obds_diamonds and obds_chickweight with labels A and B, with 2 columns and 1 row
  
super_plot <- cowplot::plot_grid(first_row, obds_msleep, labels=c("", "C"), nrow=2) # stitch the  first row plot together with the obds_sleep
super_plot #display the superplot

```

Using `patchwork`.

```{r, fig.height=12, fig.width=12}
library(patchwork)
(obds_diamonds + obds_chickweight/obds_msleep)

```

Using `gridExtra`.

```{r, fig.height=12, fig.width=12}
library(gridExtra)
gridExtra::grid.arrange(obds_diamonds, obds_chickweight, obds_msleep, ncol=2, nrow=1)




```

- Export the new plot in a PDF file, and open it in a PDF viewer (e.g. Adobe Acrobat Reader DC).

You will likely need a few attempts to fine-tune the width and height of the output file.

```{r}
#ggsave( 
  where_do_you_want_to_save,
  plot=ggplot_day4.pdf, #if you save the file with the name pdf it will save as a pdf
  width=...,
  height=...)
```

# Exercise

## Pair programming

- Explore the data set `ggplot2::mpg` and generate the most informative plot that you can!

```{r}




```

